---
title: "C5W2 assignment"
author: "DT"
date: "February 21, 2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(dplyr)
library(ggplot2)
library(lattice)

```

##Reading in the dataset, creating the "raw" file, then pulling out the "NA"
```{r load 1, 2}
raw_file <- read.csv("activity.csv")
data <- na.omit(raw_file)

```

##Daily descriptive statistics for step count
```{r daily 1, 3}
daily_sum <- aggregate(steps ~ date, data, sum)
daily_mean <- mean(daily_sum$steps)
daily_median <- median(daily_sum$steps)

```
The mean daily step count is `r daily_mean` and the daily median is `r daily_median`.

```{r daily 2}
hist(daily_sum$steps, main = "Total Steps Each Day", xlab = "Number of Steps", ylab = "Frequency", col = "orange")

```
```{r daily pattern 1}
interval_average <- aggregate(steps ~ interval, data, mean)
plot(x = interval_average$interval, y = interval_average$steps, xlab = "Interval"
     , ylab = "Average number of steps", main = "Average Daily Number of Steps by Interval", 
     type = "l")

```

```{r daily pattern 2}
fiveminmax <- max(interval_average$steps)
maxinterval <- interval_average[interval_average$steps==fiveminmax,1]

```
The highest average step count for an interval was `r fiveminmax`, which was during the `r maxinterval` interval.

##How to handle missing values
```{r missingno 1}
number_na <- sum(is.na(raw_file$steps))

```
Total number of missing values in the dataset is `r number_na`.

```{r missingno 2, 3}
na_data <- raw_file[is.na(raw_file$steps),]
non_na_data <- merge(na_data, interval_average, by = "interval")
non_na_data <- non_na_data[,c(4,3,1)]
colnames(non_na_data) <- c("steps", "date", "interval")
complete_data <- rbind(data, non_na_data)

```
Missing values were replaced with the average for that interval across the entire sampling period.


```{r missingno 4}
missingno_sum <- aggregate(steps ~ date, complete_data, sum)
missingno_mean <- mean(missingno_sum$steps)
missingno_median <- median(missingno_sum$steps)
hist(missingno_sum$steps, main = "Total Steps Each Day Including Missing Values",xlab = "Number of Steps", ylab = "Frequency", col = "orange")
complete_sum <- sum(missingno_sum$steps) 
cleaned_sum <- sum(daily_sum$steps)

```
After imputing missing values, the new mean daily step count is `r missingno_mean` and the daily median is `r missingno_median`
The values do not differ from the previous estimates - `r daily_mean` and `r daily_median` as the previous mean and median estimate, respectively, because I replaced missing values with the mean for that interval, so I wouldn't expect any differences in the basic descriptive statistics. Imputing missing data greatly increased the total daily number of steps: the imputed total (`r cleaned_sum`) is quite a bit larger than the pre-replacement total (`r complete_sum`)


##weekday vs weekend?
```{r week 1}
complete_data$week <- as.factor(weekdays(as.Date(complete_data$date)))
final_dataset <- complete_data
final_dataset$daytype <- as.factor(ifelse(final_dataset$week %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

```

```{r week 2}
final_dataset_table <- aggregate(steps ~ interval + daytype, final_dataset, mean)
with(final_dataset_table, xyplot(steps ~ interval|daytype, main = "Average Steps per Day by Interval for Weekend vs Weekdays", xlab = "Interval", ylab = "Steps", 
                                 layout = c(1, 2),type = "l"))
""

```
The activity pattern for weekdays compared to weekends are noticably different. Weekdays tend to start and end earlier, with fewer steps taken overall. 





